\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-07-25 analysis/paper/paper.Rmd
review theory and implemntation sections

# Code formatting

@@

Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative.

As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
$$\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},$$
where $N(t)$ is the counting process associated with the event of interest, $\lambda(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling. This parametric form suggests that we can readily estimate log-hazards of the form $\log(\lambda(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $\log(\rho(t))$ as an offset.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

@@

$$ h(x,t) = exp[g(x,t)] $$
where $t$ denotes the numerical value (number of units) of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any). Different functions of $t$ lead to different parametric hazard models.

The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.

$$ h(x,t) = exp(\beta_0 + \beta_1 x) $$

The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been on study a.k.a the memoryless property (Kalbfleisch and Prentice, 2002).

@@

$$ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) $$

Use of $log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

$$ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) $$

@@

The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) data analysis.

We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our testing units.

@@

## Data analysis

The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function. By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].

@@

Finally, the hazard function is fitted to the data using the function \code{glm}, unless there is more than one type event (i.e. in a competing-risk analysis), in which case it uses the multinomical regression capabilities of the \pkg{VGAM} package. This package was selected for its ability to fit multinomial regression models with an offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. For the case of a competing-event analysis, the event-specific risk is computed using a nested double integral; in this setting, Monte-Carlo integration is faster and more flexible than \code{stats::integrate}. This is due to the fact that the computation involves a double loop over the selected time points; Monte-Carlo integration performs this step by using the vectorized \code{rowSums} and \code{colSums} functions.

To decide between a single-event and a competing-event analysis, we created \code{absoluteRisk} as an \code{S3} generic, with methods for both \code{glm} and \code{CompRisk} objects (the latter inherits from \code{vglm} as well). The method dispatch system of \proglang{R} then takes care of matching the correct input to the correct methodology, without the user's intervention.

@@

$$ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$

@@

```{r erspc-casebase-splines, eval = eval_cs1}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ splines::bs(Follow.Up.Time) + ScrArm, 
                                    data = ERSPC, 
                                    ratio = 100)

@@

# Case study 3

  - Glmnet, gam, and gbm

>>>>

# Code formatting {.unnumbered}

@@

<!-- Popular survival analysis methods, like Kaplan-Meier and Cox regression, rely on the notion of risk-set sampling. Case-base sampling can be seen as an alternative. -->

As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling. This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

@@

$$ h(t;X) = \exp[g(t;X)] $$
where $t$ denotes time and $X$, the individual's covariate profile. Different functions of $t$ lead to different parametric hazard models. The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.

$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$

The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).

@@

$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$

Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$

For competing-risk analyses with $J$ possible events, we can show that each person-moment's contribution of the likelihood is of the form

$$\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},$$

where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}

@@

The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling. We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

## Parametric modeling

The parametric modeling step was separated into three parts:

  1. case-base sampling; 
  2. estimation of the smooth hazard function; 
  3. calculation of the risk function.

By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].

@@

To fit the hazard function, we provide several approaches that are available via the \code{family} parameter. These approaches are:

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net. This functionality is provided through the \code{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \code{gam} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \code{gbm} package. This feature is still experimental.

In the case of multiple events, the hazard is fitted via multinomial regression as performed by the \pkg{VGAM} package. This package was selected for its ability to fit multinomial regression models with an offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as

$$ CI(t) = 1 - S(t).$$

For the case of a competing-event analysis, the event-specific risk is computed using the following procedure: first, we compute the overall survival function (i.e. for all event types):

$$ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).$$
From this, we can derive the event-specific subdensities:

$$ f_j(t) = h_j(t)S(t).$$

Finally, by integrating these subdensities, we obtain the event-specific cumulative incidence functions:

$$ CI_j(t) = \int_0^t f_j(u)du.$$

We created \code{absoluteRisk} as an \code{S3} generic, with methods for the different types of outputs of \code{fitSmoothHazard}. The method dispatch system of \proglang{R} then takes care of matching the correct output to the correct methodology for calculating the cumulative incidence function, without the user's intervention.

In the following sections, we illustrate these functionalities in the context of three case studies.

@@

$$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$

@@

```{r erspc-casebase-splines, eval = eval_cs1}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
                                    data = ERSPC, 
                                    ratio = 100)

@@

# Case study 3--The Cancer Genome Atlas

  - Glmnet and TCGA data

### 2019-07-25 analysis/paper/paper.tex
review theory and implemntation sections

\Address{
    Sahir Bhatnagar *\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  186 Dysart Road Winnipeg, MB, Canada R3B 0S8\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{https://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

@@

\hypertarget{code-formatting}{%
\section{Code formatting}\label{code-formatting}}

@@

Popular survival analysis methods, like Kaplan-Meier and Cox regression,
rely on the notion of risk-set sampling. Case-base sampling can be seen
as an alternative.

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogenous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{\lambda(t)^{dN(t)}}{\rho(t) + \lambda(t)},\] where \(N(t)\) is
the counting process associated with the event of interest,
\(\lambda(t)\) is the corresponding hazard function, and \(\rho(t)\) is
the hazard function for the Poisson process associated with case-base
sampling. This parametric form suggests that we can readily estimate
log-hazards of the form \(\log(\lambda(t)) = g(t; X)\) using logistic
regression, where each observation corresponds to a person moment, the
function \(g(t; X)\) is linear in a finite number of parameters, and
where we treat \(\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base samping \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds to a time-homogeneous
Poisson process with intensity equal to \(b/B\), where \(b\) is the
number of sampled observations in the base series, and \(B\) is the
total population-time for the study base. More complex examples are also
available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

@@

\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
(number of units) of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any). Different functions of \(t\) lead to
different parametric hazard models.

The simplest of these models is the one-parameter exponential
distribution which is obtained by taking the hazard function to be
constant over the range of \(t\).

\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been on study a.k.a
the memoryless property (Kalbfleisch and Prentice, 2002).

@@

\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]

Use of \(log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]

@@

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

The data analysis step was separated into three parts: 1) case-base
sampling; 2) estimation of the smooth hazard function; and 3)
calculation of the risk function. By separating the sampling and
estimation functions, we allowed the possibility of users implementing
more complex sampling scheme, as described in Saarela
\citeyearpar{saarela2016case}.

@@

Finally, the hazard function is fitted to the data using the function
\code{glm}, unless there is more than one type event (i.e.~in a
competing-risk analysis), in which case it uses the multinomical
regression capabilities of the \pkg{VGAM} package. This package was
selected for its ability to fit multinomial regression models with an
offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
is computed using either the \code{stats::integrate} function or
Monte-Carlo integration. For the case of a competing-event analysis, the
event-specific risk is computed using a nested double integral; in this
setting, Monte-Carlo integration is faster and more flexible than
\code{stats::integrate}. This is due to the fact that the computation
involves a double loop over the selected time points; Monte-Carlo
integration performs this step by using the vectorized \code{rowSums}
and \code{colSums} functions.

To decide between a single-event and a competing-event analysis, we
created \code{absoluteRisk} as an \code{S3} generic, with methods for
both \code{glm} and \code{CompRisk} objects (the latter inherits from
\code{vglm} as well). The method dispatch system of \proglang{R} then
takes care of matching the correct input to the correct methodology,
without the user's intervention.

@@

\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]

@@

\begin{CodeInput}
R> library(splines)
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-3}{%
\section{Case study 3}\label{case-study-3}}

\begin{itemize}
\tightlist
\item
  Glmnet, gam, and gbm
\end{itemize}

@@

This report was generated on 2019-07-22 16:50:40 using the following
computational environment and dependencies:

@@

#>  language (EN)                        

@@

#>  date     2019-07-22                  

@@

#>  assertthat    0.2.0      2017-04-11 [1] CRAN (R 3.5.0)                  
#>  backports     1.1.2      2017-12-13 [1] CRAN (R 3.5.0)                  
#>  broom         0.4.4      2018-03-29 [1] CRAN (R 3.5.0)                  
#>  callr         3.3.0      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  casebase    * 0.1.0.9000 2019-07-19 [1] local                           
#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  

@@

#>  colorspace    1.3-2      2016-12-14 [1] CRAN (R 3.5.0)                  
#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
#>  data.table    1.11.2     2018-05-08 [1] CRAN (R 3.5.0)                  
#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  

@@

#>  dplyr       * 0.8.0.1    2019-02-15 [1] CRAN (R 3.5.2)                  

@@

#>  forcats     * 0.3.0      2018-02-19 [1] CRAN (R 3.5.0)                  
#>  foreign       0.8-70     2018-04-23 [1] CRAN (R 3.5.0)                  

@@

@@

#>  gtable        0.2.0      2016-02-26 [1] CRAN (R 3.5.0)                  
#>  haven         1.1.1      2018-01-18 [1] CRAN (R 3.5.0)                  
#>  hms           0.4.2      2018-03-10 [1] CRAN (R 3.5.0)                  
#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
#>  httr          1.3.1      2017-08-20 [1] CRAN (R 3.5.0)                  

@@

#>  lazyeval      0.2.1      2017-10-29 [1] CRAN (R 3.5.0)                  
#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  

@@

#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
#>  mnormt        1.5-5      2016-10-15 [1] CRAN (R 3.5.0)                  
#>  modelr        0.1.2      2018-05-11 [1] CRAN (R 3.5.0)                  
#>  munsell       0.4.3      2016-02-13 [1] CRAN (R 3.5.0)                  

@@

#>  pillar        1.3.1      2018-12-15 [1] CRAN (R 3.5.2)                  

@@

#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.5.2)                  

@@

#>  plyr          1.8.4      2016-06-08 [1] CRAN (R 3.5.0)                  
#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
#>  processx      3.4.0      2019-07-03 [1] CRAN (R 3.6.1)                  

@@

#>  psych         1.8.4      2018-05-06 [1] CRAN (R 3.5.0)                  
#>  purrr       * 0.2.4      2017-10-18 [1] CRAN (R 3.5.0)                  
#>  R6            2.2.2      2017-06-17 [1] CRAN (R 3.5.0)                  

@@

#>  readr       * 1.1.1      2017-05-16 [1] CRAN (R 3.5.0)                  
#>  readxl        1.1.0      2018-04-20 [1] CRAN (R 3.5.0)                  

@@

#>  reshape2      1.4.3      2017-12-11 [1] CRAN (R 3.5.0)                  

@@

#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
#>  rstudioapi    0.7        2017-09-07 [1] CRAN (R 3.5.0)                  
#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
#>  rvest         0.3.2      2016-06-17 [1] CRAN (R 3.5.0)                  
#>  scales        0.5.0      2017-08-24 [1] CRAN (R 3.5.0)                  

@@

#>  testthat      2.1.1      2019-04-23 [1] CRAN (R 3.6.1)                  
#>  tibble      * 2.0.1      2019-01-12 [1] CRAN (R 3.5.2)                  
#>  tidyr       * 0.8.1      2018-05-18 [1] CRAN (R 3.5.0)                  
#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  

@@

@@

#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  

@@

#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.5.0)                  

@@

@@

#> [1] /home/mturgeon/Rlibs

@@

\begin{CodeOutput}
#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [c8ac771] 2019-07-22: Merge pull request #5 from sahirbhatnagar/litreview
\end{CodeOutput}
\end{CodeChunk}

>>>>

\Address{
    Sahir Bhatnagar *\\
  McGill Univeristy\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{sahir.bhatnagar@mail.mcgill.ca}\\
  URL: \url{http://sahirbhatnagar.com/}\\~\\
      Maxime Turgeon *\\
  University of Manitoba\\
  318 Machray Hall Winnipeg, MB, Canada R3T 2N2\\
  E-mail: \email{max.turgeon@umanitoba.ca}\\
  URL: \url{http://maxturgeon.ca/}\\~\\
      Jesse Islam\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{jesse.islam@mail.mcgill.ca}\\

@@

\hypertarget{code-formatting}{%
\section*{Code formatting}\label{code-formatting}}
\addcontentsline{toc}{section}{Code formatting}

@@

As shown by Saarela \& Arjas \citeyearpar{saarela2015non} (and further
expanded in Saarela \citeyearpar{saarela2016case}), writing the
likelihood arising from this data-generating mechanism using the
framework of non-homogenous Poisson processes, we eventually reach an
expression where each person-moment's contribution is of the form
\[\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},\] where \(N(t)\) is the counting
process associated with the event of interest, \(h(t)\) is the
corresponding hazard function, and \(\rho(t)\) is the hazard function
for the Poisson process associated with case-base sampling. This
parametric form suggests that we can readily estimate log-hazards of the
form \(\log(h(t)) = g(t; X)\) using logistic regression, where each
observation corresponds to a person moment, the function \(g(t; X)\) is
linear in a finite number of parameters, and where we treat
\(-\log(\rho(t))\) as an offset.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest performing case-base samping \emph{uniformly}, i.e.~to sample
the base series uniformly from the study base. In terms of Poisson
processes, this sampling strategy corresponds essentially to a
time-homogeneous Poisson process with intensity equal to \(b/B\), where
\(b\) is the number of sampled observations in the base series, and
\(B\) is the total population-time for the study base. More complex
examples are also available; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the probabilities of the sampling
mechanism are proportional to the cardiovascular disease event rate
given by the Framingham score.

@@

\[ h(t;X) = \exp[g(t;X)] \] where \(t\) denotes time and \(X\), the
individual's covariate profile. Different functions of \(t\) lead to
different parametric hazard models. The simplest of these models is the
one-parameter exponential distribution which is obtained by taking the
hazard function to be constant over the range of \(t\).

\[ h(t;X) = \exp(\beta_0 + \beta_1 X) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been in the study.
This is also known as the \emph{memoryless property} (Kalbfleisch and
Prentice, 2002).

@@

\[ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) \]

Use of \(\log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) \]

For competing-risk analyses with \(J\) possible events, we can show that
each person-moment's contribution of the likelihood is of the form

\[\frac{h_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^Jh_j(t)},\]

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(h_j(t)\) is the corresponding hazard function. As may
be expected, this functional form is similar to the terms appearing in
the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}

@@

\hypertarget{parametric-modeling}{%
\subsection{Parametric modeling}\label{parametric-modeling}}

The parametric modeling step was separated into three parts:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  case-base sampling;
\item
  estimation of the smooth hazard function;
\item
  calculation of the risk function.
\end{enumerate}

By separating the sampling and estimation functions, we allowed the
possibility of users implementing more complex sampling scheme, as
described in Saarela \citeyearpar{saarela2016case}.

@@

To fit the hazard function, we provide several approaches that are
available via the \code{family} parameter. These approaches are:

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using Lasso
  or elastic-net. This functionality is provided through the
  \code{glmnet} package \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \code{gam} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \code{gbm} package. This feature is still experimental.
\end{itemize}

In the case of multiple events, the hazard is fitted via multinomial
regression as performed by the \pkg{VGAM} package. This package was
selected for its ability to fit multinomial regression models with an
offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
is computed using either the \code{stats::integrate} function or
Monte-Carlo integration. The risk function (or cumulative incidence
function) is then defined as

\[ CI(t) = 1 - S(t).\]

For the case of a competing-event analysis, the event-specific risk is
computed using the following procedure: first, we compute the overall
survival function (i.e.~for all event types):

\[ S(t) = \exp\left(-\int_0^t H(u;X) du\right),\qquad H(t;X) = \sum_{j=1}^J h_j(t;X).\]
From this, we can derive the event-specific subdensities:

\[ f_j(t) = h_j(t)S(t).\]

Finally, by integrating these subdensities, we obtain the event-specific
cumulative incidence functions:

\[ CI_j(t) = \int_0^t f_j(u)du.\]

We created \code{absoluteRisk} as an \code{S3} generic, with methods for
the different types of outputs of \code{fitSmoothHazard}. The method
dispatch system of \proglang{R} then takes care of matching the correct
output to the correct methodology for calculating the cumulative
incidence function, without the user's intervention.

In the following sections, we illustrate these functionalities in the
context of three case studies.

@@

\[ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-3the-cancer-genome-atlas}{%
\section{Case study 3--The Cancer Genome
Atlas}\label{case-study-3the-cancer-genome-atlas}}

\begin{itemize}
\tightlist
\item
  Glmnet and TCGA data
\end{itemize}

@@

This report was generated on 2019-07-25 16:51:47 using the following
computational environment and dependencies:

@@

#>  language en_CA:en                    

@@

#>  date     2019-07-25                  

@@

#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)
#>  backports     1.1.4      2019-04-10 [1] CRAN (R 3.6.1)
#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)
#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)
#>  casebase    * 0.2.1.9000 2019-07-25 [1] local         
#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.6.1)

@@

#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)
#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.6.1)
#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)
#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.6.1)

@@

#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)

@@

#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)

@@

#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)

@@

#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)
#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)
#>  hms           0.5.0      2019-07-09 [1] CRAN (R 3.6.1)
#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.6.1)
#>  httr          1.4.0      2018-12-11 [1] CRAN (R 3.6.1)

@@

#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)
#>  lubridate     1.7.4      2018-04-11 [1] CRAN (R 3.6.1)
#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.6.1)

@@

#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.6.1)
#>  mgcv          1.8-28     2019-03-21 [4] CRAN (R 3.5.3)
#>  modelr        0.1.4      2019-02-18 [1] CRAN (R 3.6.1)
#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)

@@

#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)

@@

#>  pkgconfig     2.0.2      2018-08-16 [1] CRAN (R 3.6.1)

@@

#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.6.1)
#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)

@@

#>  purrr       * 0.3.2      2019-03-15 [1] CRAN (R 3.6.1)
#>  R6            2.4.0      2019-02-14 [1] CRAN (R 3.6.1)

@@

#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)
#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)

@@

@@

#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.6.1)
#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)
#>  rticles       0.9        2019-07-12 [1] CRAN (R 3.6.1)
#>  rvest         0.3.4      2019-05-15 [1] CRAN (R 3.6.1)
#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)

@@

#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)
#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)
#>  tidyr       * 0.8.3      2019-03-01 [1] CRAN (R 3.6.1)
#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.6.1)
#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.6.1)

@@

#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)

@@

#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.6.1)

@@

#>  xml2          1.2.0      2018-01-24 [1] CRAN (R 3.6.1)

@@

#>  zeallot       0.1.0      2018-01-28 [1] CRAN (R 3.6.1)

@@

#> [1] /home/turgeonm/R/x86_64-pc-linux-gnu-library/3.6

@@

\begin{CodeOutput}
#> Local:    review-max /home/turgeonm/Documents/git_repos/cbpaper
#> Head:     [c0cb26d] 2019-07-23: Merge pull request #6 from sahirbhatnagar/fixdocker
\end{CodeOutput}
\end{CodeChunk}
